---
title: "Benchmarking BamScale Across Step1, GAlignments, and SeqQual Workloads"
format:
  html:
    toc: true
    number-sections: true
vignette: >
  %\VignetteIndexEntry{Benchmarking BamScale Across Step1, GAlignments, and SeqQual Workloads}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
params:
  step1_ga_results_dir: "run_20260320_133141"
  seqqual_results_dir: "run_20260320_162359"
execute:
  echo: false
  warning: false
  message: false
---

## Overview

This article combines the final benchmark runs used for the current BamScale benchmark summary:

- `step1` and `galignments` from `r params$step1_ga_results_dir`
- `seqqual` from `r params$seqqual_results_dir`

These runs were generated with the same server-first benchmark harness, the same balanced profile family, the same deterministic case order, and the same worker/thread budget policy. `Seqqual` is reported separately because it includes both the fair compatibility track and the optimized compact track.

## Data Loading

```{r}
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scales)
})

resolve_results_dir <- function(path) {
  input_dir <- tryCatch(dirname(normalizePath(knitr::current_input(), winslash = "/", mustWork = TRUE)), error = function(e) NA_character_)
  wd <- tryCatch(normalizePath(getwd(), winslash = "/", mustWork = TRUE), error = function(e) NA_character_)
  candidates <- c(
    path,
    if (!is.na(input_dir)) file.path(input_dir, path) else NA_character_,
    if (!is.na(input_dir)) file.path(input_dir, "..", path) else NA_character_,
    if (!is.na(input_dir)) file.path(input_dir, "..", "..", path) else NA_character_,
    if (!is.na(input_dir)) file.path(input_dir, "..", "inst", "benchmarks", path) else NA_character_,
    if (!is.na(wd)) file.path(wd, path) else NA_character_,
    if (!is.na(wd)) file.path(wd, "inst", "benchmarks", path) else NA_character_
  )
  candidates <- unique(candidates[!is.na(candidates) & nzchar(candidates)])
  hits <- candidates[file.exists(candidates)]
  if (length(hits) == 0) {
    stop("Could not locate results_dir: ", path, call. = FALSE)
  }
  normalizePath(hits[[1]], winslash = "/", mustWork = TRUE)
}

read_if_exists <- function(path) {
  if (file.exists(path)) readr::read_csv(path, show_col_types = FALSE) else tibble::tibble()
}

load_run <- function(path_label) {
  dir_path <- resolve_results_dir(path_label)
  summary_path <- if (file.exists(file.path(dir_path, "summary_qmd.csv"))) {
    file.path(dir_path, "summary_qmd.csv")
  } else {
    file.path(dir_path, "summary.csv")
  }

  list(
    label = basename(dir_path),
    dir = dir_path,
    summary = read_if_exists(summary_path),
    files = read_if_exists(file.path(dir_path, "files.csv")),
    host = read_if_exists(file.path(dir_path, "host_info.csv")),
    reference_counts = read_if_exists(file.path(dir_path, "reference_counts.csv")),
    correctness = read_if_exists(file.path(dir_path, "correctness_preflight.csv"))
  )
}

run_main <- load_run(params$step1_ga_results_dir)
run_seqqual <- load_run(params$seqqual_results_dir)

main_ok <- run_main$summary %>%
  filter(status == "ok", workload %in% c("step1", "galignments"), comparison_track == "fair")

seq_ok <- run_seqqual$summary %>%
  filter(status == "ok", workload == "seqqual")

single_main <- main_ok %>% filter(scenario == "single")
multi_main <- main_ok %>% filter(scenario == "multi")

seq_single <- seq_ok %>% filter(scenario == "single")
seq_multi <- seq_ok %>% filter(scenario == "multi")

host_main <- if (nrow(run_main$host) > 0) run_main$host[1, ] else tibble::tibble()
host_seq <- if (nrow(run_seqqual$host) > 0) run_seqqual$host[1, ] else tibble::tibble()
```

## Benchmark Provenance

- Step1 and GAlignments run: `r run_main$label`
- SeqQual run: `r run_seqqual$label`
- Step1/GAlignments results directory: `r run_main$dir`
- SeqQual results directory: `r run_seqqual$dir`

```{r}
run_overview_tbl <- tibble::tibble(
  run = c(run_main$label, run_seqqual$label),
  workloads = c("step1, galignments", "seqqual"),
  profile = c(
    if (nrow(run_main$host) > 0) host_main$profile[[1]] else NA_character_,
    if (nrow(run_seqqual$host) > 0) host_seq$profile[[1]] else NA_character_
  ),
  cpu = c(
    if (nrow(run_main$host) > 0) host_main$cpu_model[[1]] else NA_character_,
    if (nrow(run_seqqual$host) > 0) host_seq$cpu_model[[1]] else NA_character_
  ),
  logical_cores = c(
    if (nrow(run_main$host) > 0) host_main$logical_cores[[1]] else NA_integer_,
    if (nrow(run_seqqual$host) > 0) host_seq$logical_cores[[1]] else NA_integer_
  ),
  total_mem_gb = c(
    if (nrow(run_main$host) > 0) round(host_main$total_mem_gb[[1]], 1) else NA_real_,
    if (nrow(run_seqqual$host) > 0) round(host_seq$total_mem_gb[[1]], 1) else NA_real_
  ),
  successful_cases = c(nrow(main_ok), nrow(seq_ok))
)

knitr::kable(run_overview_tbl, col.names = c("Run", "Workloads", "Profile", "CPU", "Logical cores", "RAM (GB)", "Successful cases"))
```

## Methods Rationale

This benchmark suite covers three distinct access patterns:

- `step1`: alignment-metadata extraction without sequence/quality materialization. This is a good proxy for BAM filtering, fragment-length profiling, and fragment-distribution QC.
- `galignments`: construction of alignment objects suitable for Bioconductor workflows centered on `GenomicAlignments`.
- `seqqual`: full sequence and quality extraction. This is reported in two BamScale modes:
  - `fair` compatibility mode, which is the closest comparator to `Rsamtools::scanBam`
  - `optimized` compact mode, which reduces internal overhead and should be interpreted as an optimized BamScale track rather than a strict apples-to-apples replacement for the compatibility path

Across all runs, the benchmark design emphasizes:

- deterministic case order
- balanced BamScale worker/thread plans
- explicit comparator baselines
- single-file and multi-file reporting
- fixed 48-thread budget for multi-file plans

## Input Files

The same underlying four selected BAMs were used for both runs, with repeated files allowed to populate the 12-file multi-file benchmark set.

```{r}
run_main$files %>%
  transmute(
    file,
    source,
    selected_for_single,
    selected_for_multi,
    size_mb = round(size_mb, 1),
    has_index
  ) %>%
  knitr::kable()
```

## Reference Counts

```{r}
bind_rows(
  run_main$reference_counts %>% mutate(run = run_main$label),
  run_seqqual$reference_counts %>% mutate(run = run_seqqual$label)
) %>%
  mutate(total_mb = round(total_mb, 1)) %>%
  select(run, scenario, workload, n_files, n_records, total_mb) %>%
  knitr::kable()
```

## Best-Observed Summary

```{r}
best_summary_tbl <- bind_rows(main_ok, seq_ok) %>%
  group_by(scenario, workload, comparison_track, method_family) %>%
  slice_min(order_by = median_s, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    plan = paste0(bp_workers_effective, "x", threads_effective),
    median_s = round(median_s, 3)
  ) %>%
  select(scenario, workload, comparison_track, method_family, method, plan, median_s)

knitr::kable(best_summary_tbl, col.names = c("Scenario", "Workload", "Track", "Method family", "Method", "Plan", "Median time (s)"))
```

## BamScale-versus-Comparator Fold Change

```{r}
best_fold_tbl <- function(df, scenario_name, include_optimized = FALSE) {
  bamscale_fair <- df %>%
    filter(scenario == scenario_name, method_family == "BamScale", comparison_track == "fair") %>%
    group_by(workload) %>%
    slice_min(order_by = median_s, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(
      scenario = scenario_name,
      workload,
      track = "fair",
      bamscale_method = method,
      bamscale_plan = paste0(bp_workers_effective, "x", threads_effective),
      bamscale_s = median_s
    ) %>%
    left_join(
      df %>%
        filter(scenario == scenario_name, method_family != "BamScale", comparison_track == "fair") %>%
        group_by(workload) %>%
        slice_min(order_by = median_s, n = 1, with_ties = FALSE) %>%
        ungroup() %>%
        transmute(
          workload,
          comparator_method = method,
          comparator_plan = paste0(bp_workers_effective, "x", threads_effective),
          comparator_s = median_s
        ),
      by = "workload"
    )

  if (!include_optimized) {
    return(bamscale_fair)
  }

  bamscale_opt <- df %>%
    filter(scenario == scenario_name, method_family == "BamScale", comparison_track == "optimized") %>%
    group_by(workload) %>%
    slice_min(order_by = median_s, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    transmute(
      scenario = scenario_name,
      workload,
      track = "optimized",
      bamscale_method = method,
      bamscale_plan = paste0(bp_workers_effective, "x", threads_effective),
      bamscale_s = median_s
    ) %>%
    left_join(
      df %>%
        filter(scenario == scenario_name, method_family != "BamScale", comparison_track == "fair") %>%
        group_by(workload) %>%
        slice_min(order_by = median_s, n = 1, with_ties = FALSE) %>%
        ungroup() %>%
        transmute(
          workload,
          comparator_method = method,
          comparator_plan = paste0(bp_workers_effective, "x", threads_effective),
          comparator_s = median_s
        ),
      by = "workload"
    )

  bind_rows(bamscale_fair, bamscale_opt)
}

comparison_tbl <- bind_rows(
  best_fold_tbl(main_ok, "single", include_optimized = FALSE),
  best_fold_tbl(main_ok, "multi", include_optimized = FALSE),
  best_fold_tbl(seq_ok, "single", include_optimized = TRUE),
  best_fold_tbl(seq_ok, "multi", include_optimized = TRUE)
) %>%
  mutate(
    comparator_over_bamscale = comparator_s / bamscale_s,
    faster_pct = (comparator_s - bamscale_s) / comparator_s * 100,
    bamscale_s = round(bamscale_s, 3),
    comparator_s = round(comparator_s, 3),
    comparator_over_bamscale = round(comparator_over_bamscale, 3),
    faster_pct = round(faster_pct, 1)
  ) %>%
  arrange(factor(scenario, levels = c("single", "multi")), workload, track)

knitr::kable(
  comparison_tbl,
  col.names = c(
    "Scenario",
    "Workload",
    "Track",
    "BamScale method",
    "BamScale plan",
    "BamScale (s)",
    "Comparator method",
    "Comparator plan",
    "Comparator (s)",
    "Comparator / BamScale",
    "BamScale faster (%)"
  )
)
```

```{r}
comparison_plot_df <- comparison_tbl %>%
  mutate(
    label = ifelse(track == "fair", paste(scenario, workload, sep = " | "), paste(scenario, workload, "compact", sep = " | ")),
    label = factor(label, levels = label)
  )

ggplot(comparison_plot_df, aes(x = label, y = comparator_over_bamscale, fill = track)) +
  geom_col(width = 0.72) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
  geom_text(aes(label = sprintf("%.2fx", comparator_over_bamscale)), vjust = -0.2, size = 3.4) +
  scale_fill_manual(values = c(fair = "#4C78A8", optimized = "#C44E52")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Best-observed BamScale fold change versus comparator",
    subtitle = "Values above 1 indicate faster BamScale runtime",
    x = NULL,
    y = "Comparator time / BamScale time",
    fill = "Track"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

## Step1 and GAlignments: Single-File Scaling

```{r}
bamscale_single_main <- single_main %>% filter(method_family == "BamScale")

ggplot(bamscale_single_main, aes(x = threads_effective, y = median_s, color = workload)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.6) +
  scale_x_continuous(breaks = sort(unique(bamscale_single_main$threads_effective))) +
  scale_color_manual(values = c(step1 = "#C44E52", galignments = "#4C78A8")) +
  labs(
    title = "BamScale single-file scaling",
    subtitle = "Step1 and GAlignments",
    x = "Threads",
    y = "Median elapsed time (s)",
    color = "Workload"
  ) +
  theme_minimal(base_size = 12)
```

```{r}
single_compare_main <- single_main %>%
  group_by(workload, method_family) %>%
  slice_min(order_by = median_s, n = 1, with_ties = FALSE) %>%
  ungroup()

ggplot(single_compare_main, aes(x = method, y = median_s, fill = method_family)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.2f", median_s)), vjust = -0.2, size = 3.4) +
  facet_wrap(~workload, scales = "free_y") +
  scale_fill_manual(values = c(BamScale = "#C44E52", Rsamtools = "#4C78A8", GenomicAlignments = "#59A14F")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Best single-file runtime by method family",
    x = NULL,
    y = "Median elapsed time (s)",
    fill = "Method family"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 16, hjust = 1))
```

## Step1: Multi-File Scaling

```{r}
multi_step1 <- multi_main %>% filter(workload == "step1")

ggplot(multi_step1, aes(x = bp_workers_effective, y = median_s, color = method_family)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.6) +
  scale_x_continuous(breaks = sort(unique(multi_step1$bp_workers_effective))) +
  scale_color_manual(values = c(BamScale = "#C44E52", Rsamtools = "#4C78A8", GenomicAlignments = "#59A14F")) +
  labs(
    title = "Multi-file step1 scaling",
    subtitle = "Balanced BamScale plans versus Rsamtools + BiocParallel",
    x = "Workers",
    y = "Median elapsed time (s)",
    color = "Method family"
  ) +
  theme_minimal(base_size = 12)
```

## GAlignments: Multi-File Scaling

```{r}
multi_ga <- multi_main %>% filter(workload == "galignments")

ggplot(multi_ga, aes(x = bp_workers_effective, y = median_s, color = method_family)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.6) +
  scale_x_continuous(breaks = sort(unique(multi_ga$bp_workers_effective))) +
  scale_color_manual(values = c(BamScale = "#C44E52", Rsamtools = "#4C78A8", GenomicAlignments = "#59A14F")) +
  labs(
    title = "Multi-file galignments scaling",
    subtitle = "Balanced BamScale plans versus GenomicAlignments + BiocParallel",
    x = "Workers",
    y = "Median elapsed time (s)",
    color = "Method family"
  ) +
  theme_minimal(base_size = 12)
```

## BamScale Worker-Thread Tradeoff

```{r}
bamscale_tradeoff <- multi_main %>%
  filter(method_family == "BamScale")

ggplot(bamscale_tradeoff, aes(x = bp_workers_effective, y = median_s, color = workload)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  geom_text(aes(label = paste0(bp_workers_effective, "x", threads_effective)), nudge_y = 3.5, size = 3.1, show.legend = FALSE) +
  scale_x_continuous(breaks = sort(unique(bamscale_tradeoff$bp_workers_effective))) +
  scale_color_manual(values = c(step1 = "#C44E52", galignments = "#4C78A8")) +
  labs(
    title = "BamScale worker-thread tradeoff",
    subtitle = "Balanced plans under a fixed 48-thread budget",
    x = "Workers",
    y = "Median elapsed time (s)",
    color = "Workload"
  ) +
  theme_minimal(base_size = 12)
```

## SeqQual: Single-File Fair and Compact Tracks

```{r}
seq_single_plot_df <- seq_single %>%
  mutate(
    track_label = case_when(
      comparison_track == "optimized" ~ "BamScale compact",
      method_family == "BamScale" ~ "BamScale compatible",
      TRUE ~ "Rsamtools"
    )
  )

ggplot(seq_single_plot_df, aes(x = threads_effective, y = median_s, color = track_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.6) +
  scale_x_continuous(breaks = sort(unique(seq_single_plot_df$threads_effective))) +
  scale_color_manual(values = c("BamScale compatible" = "#4C78A8", "BamScale compact" = "#C44E52", "Rsamtools" = "#59A14F")) +
  labs(
    title = "Single-file seqqual scaling",
    subtitle = "Fair compatibility track and optimized compact track",
    x = "Threads",
    y = "Median elapsed time (s)",
    color = NULL
  ) +
  theme_minimal(base_size = 12)
```

## SeqQual: Multi-File Fair and Compact Tracks

```{r}
seq_multi_plot_df <- seq_multi %>%
  mutate(
    track_label = case_when(
      comparison_track == "optimized" ~ "BamScale compact",
      method_family == "BamScale" ~ "BamScale compatible",
      TRUE ~ "Rsamtools"
    )
  )

ggplot(seq_multi_plot_df, aes(x = bp_workers_effective, y = median_s, color = track_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.6) +
  scale_x_continuous(breaks = sort(unique(seq_multi_plot_df$bp_workers_effective))) +
  scale_color_manual(values = c("BamScale compatible" = "#4C78A8", "BamScale compact" = "#C44E52", "Rsamtools" = "#59A14F")) +
  labs(
    title = "Multi-file seqqual scaling",
    subtitle = "Fair compatibility track and optimized compact track",
    x = "Workers",
    y = "Median elapsed time (s)",
    color = NULL
  ) +
  theme_minimal(base_size = 12)
```

## SeqQual Compact Gain

```{r}
seq_compact_gain_tbl <- seq_ok %>%
  filter(method_family == "BamScale") %>%
  select(scenario, comparison_track, bp_workers_effective, threads_effective, median_s) %>%
  mutate(plan = paste0(bp_workers_effective, "x", threads_effective)) %>%
  select(-bp_workers_effective, -threads_effective) %>%
  pivot_wider(names_from = comparison_track, values_from = median_s) %>%
  mutate(
    compatible = round(fair, 3),
    compact = round(optimized, 3),
    compatible_over_compact = round(fair / optimized, 3)
  ) %>%
  arrange(factor(scenario, levels = c("single", "multi")), plan) %>%
  select(scenario, plan, compatible, compact, compatible_over_compact)

knitr::kable(seq_compact_gain_tbl, col.names = c("Scenario", "Plan", "Compatible (s)", "Compact (s)", "Compatible / compact"))
```

## Benchmark Interpretation

This combined benchmark supports the following conclusions:

1. BamScale showed clear single-file and best-case multi-file advantages for both `step1` and `galignments`, with the strongest multi-file points occurring in the low-worker, high-thread regime.
2. For `seqqual`, BamScale’s fair compatibility track remained slower than the Rsamtools comparator, but the optimized compact track substantially reduced that gap and produced a clear single-file win.
3. In multi-file `seqqual`, the compact track improved on the fair compatibility path but did not surpass the best Rsamtools multi-file baseline in this run.
4. Overall, `step1` and `galignments` are the clearest BamScale strengths, while `seqqual` compact mode represents a substantial optimization that improves, but does not fully close, the multi-file gap to the comparator.
